---
title: "Deconvolution Analysis with CIBERSORT"
author: Cristiane Esteves, Mariana Boroni - Bioinformatics and Computational Biology Lab (LBBC/INCA-RJ)
data: 2022-08-01
output:
  html_document:
    df_print: paged
---

```{r}

#Load libPaths
.libPaths(c("~/deconv_cibersort/deconv_cibersort/lib/","/home/manager/R/x86_64-pc-linux-gnu-library/4.2"))
pkgs <- c("survival", "survminer", "data.table", "dplyr", "ggplot2", "e1071", "parallel", "preprocessCore", "corrplot", "RColorBrewer", "parallel", "ggdendro", "tibble" )
#install.packages(pkgs)

```

```{r}
#Load Packages
suppressPackageStartupMessages({
  library(tibble)
  library(dplyr)
  library(ggplot2)
  library(survival)
  library(survminer)
  library(e1071)
  library(parallel)
  library(preprocessCore)
  library(data.table)
  library(RColorBrewer)
  library(readr)
})

#Load script CIBERSORT and barplot functions
source('~/RNA deconvolution/Data_Deconvolution/deconv_cibersort/CIBERSORT.R')
source('~/RNA deconvolution/Data_Deconvolution/deconv_cibersort/barplot_cibersort.R')

```
#Deconvolution Analysis - CIBERSORT

CIBERSORT is a deconvolution-based method that uses the v-support vector regression method (v-SVR) to estimate each immune cell type’s relative proportion from a gene expression profile. It requires an input of reference gene expression signatures. CIBERSORT provides a gene signature matrix (LM22.txt)(https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4739640/). The file contains expression counts for 547 signature genes (547 rows) for 22 distinct human immune cells (22 columns).  

The CIBERSORT source code and LM22 signature matrix are only freely available to academic users – register for access on the CIBERSORT website .

```{r}
#load the LM22 signatures
lm22_signatures <- as.data.frame(fread("~/RNA deconvolution/Data_Deconvolution/deconv_cibersort/data/lm22.txt"))
lm22_signatures <- tibble::column_to_rownames(lm22_signatures, "V1")
#Print the first six rows of LM22 signatures dataframe
print(head(lm22_signatures[,1:4]))
#Print cell types present in the Lm22 signature
colnames(lm22_signatures)
```

# In this exercise, we will predict cell types in Bulk RNA matrix (SKCM-Metastasis)

As input, we will use TCGA-SKCM metastatic expression dataset as input. The file provided for this exercise have already been filtered to only contain genes that will be used by CIBERSORT.
Obs.: The gene names must always be in the rownames, and it should match the reference. 

```{r}

#Load the Bulk RNA TCGA-SKCM metastatic expression dataframe 
#This dataframe has been alread filtered to only contain genes from lm22 signature that will be used by cibersort)
skcm_bulk <- as.data.frame(fread("~/RNA deconvolution/Data_Deconvolution/deconv_cibersort/data/bulk.txt"))
skcm_bulk <- tibble::column_to_rownames(skcm_bulk, "V1")
#Here we printing the first six rows of Bulk RNA TCGA-SKCM dataframe
print(head(skcm_bulk[,1:4]))
```
CIBERSORT parameters:
i)   perm = No. permutations; set to >=100 to calculate p-values (default = 0).
ii)  QN = Quantile normalization of input Patient_ID (default = TRUE) - (disabling is recommended for RNA-Seq data).
iii) absolute = Run CIBERSORT in absolute mode (default = FALSE). #The CIBERSORT absolute mode scales cellular fractions to a score that reflects each cell type’s absolute proportion, allowing comparison across both samples and cell types.
                 
```{r}
# Here we are setting the seed to maintain the reproducibility of the results.
set.seed(42)
h1 <- Sys.time()
#Run CIBERSORT
#results.cibersort <- CIBERSORT(lm22_signatures, skcm_bulk, perm = 1, absolute = T, QN = F) #it is important to always use 100 or more permutations ("perm" parameter) but for didactic purposes, we will use perm = 1.
h2 <- Sys.time()
print(h2 - h1)

#Filter for the significant predictions
#results.sign = as.data.frame(results.cibersort)[which(as.data.frame(results.cibersort)$`P-value` <= 0.05),]
#results.sign = results.sign[1:22]
```

```{r}
#for a interest of time, we will upload a pre-computed result
results.sign<- readRDS("~/RNA deconvolution/Data_Deconvolution/deconv_cibersort/results_cibersort_perm100.rds")

```

```{r}
library(readr)

# Load Metastatic Melanoma (SKCM-TCGA) Clinical and Survival dataset

#data_SKCM a list format object that contains a dataframe of clinical and survival data
data_SKCM = readRDS("~/RNA deconvolution/Data_Deconvolution/deconv_cibersort/data/data_SKCM.rds")
head(data_SKCM$survival_met)

#Load the informations of the type molecular of tumor for each patient
subtypes <- read_csv("~/RNA deconvolution/Data_Deconvolution/deconv_cibersort/data/subtipos.csv", show_col_types = F)
head(subtypes)
```


```{r}
#We need unique patient IDs  in rownames (in this dataset there is more than one sample per patient).
results.sign$Patient_ID <- rownames(results.sign)

#the gsub function will help us to replace the existing dots (.) in patient IDs to a dash (-).
results.sign$Patient_ID <- gsub("\\.", "-", results.sign$Patient_ID)
#Keep the first 12 characters of the patient ID.
results.sign$Patient_ID <- substr(results.sign$Patient_ID,1,12)

#Aggregate Cibersort result with clinical and survival data according to patient ID
forest_data <- left_join(results.sign,data_SKCM$survival_met[,c(1,8,16,17,2,5)], by= c("Patient_ID" = "bcr_patient_barcode"))
forest_data <- left_join(forest_data,subtypes[,c(2,10)], by= c("Patient_ID" = "pan.samplesID"))

#rename column to "Subtype_Mutation"
colnames(forest_data)[29] = "Subtype_Mutation"
```



# Barplot
### Plot Proportions of cell types by sample predicted by Cibersort

```{r fig.height=10, fig.width=15}

# Check the columns (variable names) present in the clinical dataset
names(forest_data)

## Lets create annotation data that will be used to do the plot
#Filter for columns sample and Subtype mutation
data_barplot = forest_data[,c(23,29)]

#Make patient IDs unique values
data_barplot$Patient_ID <- make.names(data_barplot$Patient_ID, unique = T)
data_barplot$Patient_ID <- gsub("\\.", "-", data_barplot$Patient_ID)

# Put patient IDs in rownames
rownames(data_barplot) = data_barplot$Patient_ID
data_barplot$Patient_ID = NULL

# Add `NA` in empty fields
data_barplot$Subtype_Mutation[which(is.na(data_barplot$Subtype_Mutation))] <- "nan"
data_barplot$Subtype_Mutation[which(data_barplot$Subtype_Mutation == "-")] <- "nan"

# Make Patient_ID column (Patient IDs from cibersort result table) as first column
res_cibersort = forest_data[, c("Patient_ID", colnames(forest_data)[1:22])]
res_cibersort$Patient_ID <- make.names(res_cibersort$Patient_ID, unique = T)
res_cibersort$Patient_ID <- gsub("\\.", "-", res_cibersort$Patient_ID)

#Plot the barplot: in which each column we have  a patient with the clinical informations on the first row (Subtype mutation, for example, colored according to legend colors) and each bar is divided by the proportion o immune cells types predicted by cibersort. 

plot.ciber.heat(ciber.obj = res_cibersort, ann_info = data_barplot, sample.column = 1)
```


## Univariate and Multivariate (Cox Regression)/Survival analysis

The Cox proportional-hazards model (Cox, 1972) is essentially a regression model commonly used statistical in medical research for investigating the association between the survival time of patients and one or more predictor variables.

In clinical investigations, there are many situations, where several known quantities (known as covariates), potentially affect patient prognosis.

For instance, suppose two groups of patients are compared: those with and those without a specific genotype. If one of the groups also contains older individuals, any difference in survival may be attributable to genotype or age or indeed both. Hence, when investigating survival in relation to any one factor, it is often desirable to adjust for the impact of others.

Statistical model is a frequently used tool that allows to analyze survival with respect to several factors simultaneously. Additionally, statistical model provides the effect size for each factor.

The cox proportional-hazards model is one of the most important methods used for modelling survival analysis data. The next section introduces the basics of the Cox regression model.

References: 
http://www.sthda.com/english/wiki/cox-proportional-hazards-model

https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2394262/



```{r fig.height=10, fig.width=8}
library(dplyr)
library(survival)


# Univariate Cox
#Cox univariate analysis estimated the impact on survival of each cell type.
#For this, it is necessary to have the survival time of each patient and the event variable (in this case, death) and provide this information as input to the surv() function.
surv_object <- Surv(time = forest_data$OS.time, event = forest_data$OS)

#the gsub function will help us to replace the existing dots (.) in patient IDs to a dash (-).
colnames(forest_data)[1:22] <- gsub(" ", "_", colnames(forest_data)[1:22])

#Rename column from position 9 as some plots complain when the column name has dashes or dots.
colnames(forest_data)[9] <- "Treg"

# Get names of each column (immune cells and clinical variables).
covariables <- colnames(forest_data)[c(1:22,27:29)]

#Function to aggregate the column names that will be evaluated in the survival analysis by the surv() and coxph() function.
univ_formulas <- sapply(covariables, function(x) as.formula(paste('surv_object ~', x)))

#lapply is a function that will execute the surv() function (survival analysis) for each covariables obtained and store it in a list.
univ_models <- lapply(univ_formulas, function(x){coxph(x, data = forest_data)})

#Rescue the results of the previous function (surv() - survival analysis): p-value, hazard ratios and confidence intervals.
univ_results <- lapply(univ_models,
                              function(x){ 
                                x <- summary(x)
                                p.value<-signif(x$wald["pvalue"], digits=2)
                                wald.test<-signif(x$wald["test"], digits=2)
                                beta<-signif(x$coef[1], digits=2);#coeficient beta
                                HR <-signif(x$coef[2], digits=2);#exp(beta)
                                HR.confint.lower <- signif(x$conf.int[,"lower .95"], 2)
                                HR.confint.upper <- signif(x$conf.int[,"upper .95"],2)
                                HR <- paste0(HR, " (", 
                                             HR.confint.lower, "-", HR.confint.upper, ")")
                                res<-c(beta, HR, wald.test, p.value)
                                names(res)<-c("beta", "HR (95% CI for HR)", "wald.test", 
                                              "p.value")
                                return(res)
                              })

#Transform the list (univ_results) object into dataframe
res_cox = as.data.frame(t(do.call(cbind, univ_results)))
res_cox <- as.data.frame(res_cox)

#Transform pvalue column into character and numeric
res_cox$p.value <- as.character(res_cox$p.value)
res_cox$p.value <- as.numeric(res_cox$p.value)

#Filter for pval =< 0.05 as significative
res_cox_filt <- res_cox[which(res_cox$p.value <= 0.05),]
res_cox_filt

#Check the immune cells that significantly impact each patient's survival (p val =<0.05)
rownames(res_cox_filt)


```

#Multivariate Analysis

```{r}

#Now, we want to describe how the factors jointly impact on survival. To answer to this question, we’ll perform an statistical test called multivariate Cox regression analysis.

f1 <- as.formula(paste("Surv(forest_data$OS.time, event = forest_data$OS) ~ ",
                       paste(c(rownames(res_cox_filt)), collapse= "+")))
fit.coxph <- coxph(f1, data = forest_data)

summary(fit.coxph)

ggforest(fit.coxph, data = forest_data, main = "Hazard Ratio Melanoma Metastasis")
```